---
title: "Stevia RNA-seq pathogen detection report"
author:
  - name: "DFSM"
    affiliation: |
      Bioinformatics department
      NIDTec - Polytechnic School
      National University of Asuncion
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
params:
  project_dir: "."
  sample: "V350344566_L01_SE01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(jsonlite)
library(janitor)
library(knitr)
library(DT)
library(scales)
library(patchwork)

proj <- normalizePath(params$project_dir)
sample <- params$sample

read_kraken_report <- function(path) {
  readr::read_tsv(
    path,
    col_names = c("pct", "clade_reads", "taxon_reads", "rank", "taxid", "name"),
    show_col_types = FALSE
  ) %>%
    mutate(name = stringr::str_trim(name))
}

safe_read_tsv <- function(path, ...) {
  if (!file.exists(path)) {
    stop(sprintf("Required file not found: %s", path))
  }
  readr::read_tsv(path, ...)
}

safe_read_lines <- function(path) {
  if (!file.exists(path)) {
    stop(sprintf("Required file not found: %s", path))
  }
  readLines(path)
}

fastp_json_path    <- file.path(proj, "results", "trim", paste0(sample, ".fastp.json"))
hisat2_sum_path    <- file.path(proj, "results", "host", paste0(sample, ".hisat2.summary.txt"))
kraken_report_path <- file.path(proj, "results", "pathogen", paste0(sample, ".k2.report"))
bracken_sp_path    <- file.path(proj, "results", "pathogen", paste0(sample, ".bracken.species.tsv"))
bracken_gn_path    <- file.path(proj, "results", "pathogen", paste0(sample, ".bracken.genus.tsv"))
idxstats_path      <- file.path(proj, "results", "pathogen", paste0(sample, ".idxstats.tsv"))
coverage_path      <- file.path(proj, "results", "pathogen", paste0(sample, ".coverage.tsv"))
depth_path         <- file.path(proj, "results", "pathogen", paste0(sample, ".depth_top.tsv"))
blast_hits_path    <- file.path(proj, "results", "pathogen", paste0(sample, ".contigs_vs_pathogen_refs.tsv"))
refmap_path        <- file.path(proj, "refs", "validation", "reference_map.tsv")
watchlist_path     <- file.path(proj, "refs", "custom_lists", "stevia_priority_watchlist.tsv")

fastp_json <- jsonlite::fromJSON(fastp_json_path)
hisat2_summary_lines <- safe_read_lines(hisat2_sum_path)
kraken <- read_kraken_report(kraken_report_path)

bracken_sp <- safe_read_tsv(bracken_sp_path, show_col_types = FALSE)
bracken_gn <- safe_read_tsv(bracken_gn_path, show_col_types = FALSE)

idxstats <- safe_read_tsv(
  idxstats_path,
  col_names = c("ref", "length", "mapped", "unmapped"),
  show_col_types = FALSE
)

coverage <- safe_read_tsv(
  coverage_path,
  show_col_types = FALSE
) %>%
  rename(ref = 1)

depth <- if (file.exists(depth_path) && file.info(depth_path)$size > 0) {
  safe_read_tsv(
    depth_path,
    col_names = c("ref", "pos", "depth"),
    show_col_types = FALSE
  )
} else {
  tibble(ref = character(), pos = integer(), depth = integer())
}

blast_hits <- safe_read_tsv(
  blast_hits_path,
  col_names = c(
    "qseqid", "sseqid", "pident", "length", "mismatch", "gapopen",
    "qstart", "qend", "sstart", "send", "evalue", "bitscore"
  ),
  show_col_types = FALSE
)

refmap <- safe_read_tsv(
  refmap_path,
  col_names = c("ref", "header"),
  show_col_types = FALSE
)

watchlist <- safe_read_tsv(
  watchlist_path,
  show_col_types = FALSE
)

summary_tbl <- tibble(
  sample = sample,
  raw_reads = fastp_json$summary$before_filtering$total_reads,
  trimmed_reads = fastp_json$summary$after_filtering$total_reads,
  raw_bases = fastp_json$summary$before_filtering$total_bases,
  trimmed_bases = fastp_json$summary$after_filtering$total_bases,
  q20_rate_after = fastp_json$summary$after_filtering$q20_rate,
  q30_rate_after = fastp_json$summary$after_filtering$q30_rate,
  gc_content_after = fastp_json$summary$after_filtering$gc_content
)

species_tbl <- bracken_sp %>%
  arrange(desc(new_est_reads))

genus_tbl <- bracken_gn %>%
  arrange(desc(new_est_reads))

priority_hits <- species_tbl %>%
  mutate(name_lc = str_to_lower(name)) %>%
  rowwise() %>%
  mutate(
    stevia_priority = any(str_detect(name_lc, str_to_lower(watchlist$taxon))),
    priority_match = paste(
      watchlist$taxon[str_detect(name_lc, str_to_lower(watchlist$taxon))],
      collapse = "; "
    )
  ) %>%
  ungroup() %>%
  filter(stevia_priority) %>%
  arrange(desc(new_est_reads))

ref_cov <- idxstats %>%
  filter(ref != "*", mapped > 0) %>%
  left_join(refmap, by = "ref") %>%
  left_join(
    coverage %>% select(ref, numreads, covbases, coverage, meandepth),
    by = "ref"
  ) %>%
  arrange(desc(mapped))

best_contig_hits <- blast_hits %>%
  group_by(qseqid) %>%
  slice_max(order_by = bitscore, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(bitscore))

top_species_plot_tbl <- species_tbl %>%
  slice_max(order_by = new_est_reads, n = 15, with_ties = FALSE) %>%
  mutate(name = forcats::fct_reorder(name, new_est_reads))

top_refs <- ref_cov %>%
  slice_max(order_by = mapped, n = 9, with_ties = FALSE) %>%
  pull(ref)
```

## 1. Overview

This report summarizes pathogen detection from one **Stevia rebaudiana** RNA-seq sample after:

1. read QC and trimming,
2. host depletion with HISAT2,
3. taxonomic classification of non-host reads with Kraken2 and Bracken,
4. reference-level validation with Bowtie2 coverage,
5. contig assembly with rnaSPAdes and BLAST support.

## 2. Sample summary

```{r chunk2}
kable(summary_tbl, digits = 4)
```

## 3. Host depletion summary

```{r results='asis'}
cat("```\n")
cat(paste(hisat2_summary_lines, collapse = "\n"))
cat("\n```")
```

## 4. Kraken2 classification overview

```{r chunk4}
kraken_summary <- kraken %>%
  filter(rank %in% c("D", "K", "P", "C", "O", "F", "G", "S")) %>%
  arrange(desc(taxon_reads)) %>%
  head(20)

datatable(kraken_summary, options = list(pageLength = 20))
```

## 5. Bracken species-level abundance

```{r chunk5}
datatable(species_tbl, options = list(pageLength = 20, scrollX = TRUE))
```

```{r chunk5a, fig.width=10, fig.height=6}
ggplot(top_species_plot_tbl, aes(x = name, y = new_est_reads)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top detected species by Bracken estimated reads",
    x = "Species",
    y = "Estimated reads"
  ) +
  theme_bw(base_size = 11)
```

## 6. Bracken genus-level abundance

```{r chunk6}
datatable(genus_tbl, options = list(pageLength = 20, scrollX = TRUE))
```

## 7. Stevia-priority taxa

```{r chunk7}
if (nrow(priority_hits) == 0) {
  cat("No taxa from the Stevia-priority watchlist were detected in the Bracken species table.")
} else {
  datatable(priority_hits, options = list(pageLength = 15, scrollX = TRUE))
}
```

## 8. Reference-level mapping support

```{r chunk8}
if (nrow(ref_cov) == 0) {
  cat("No mapped pathogen references were found in the validation BAM.")
} else {
  datatable(ref_cov, options = list(pageLength = 20, scrollX = TRUE))
}

```

## 8b. Plot diagnostics

```{r chunk8b} 
cat("Rows in ref_cov:", nrow(ref_cov), "\n")
cat("Rows in depth:", nrow(depth), "\n")
cat("Number of top_refs:", length(top_refs), "\n")

if (length(top_refs) > 0) {
  print(head(top_refs, 10))
}

depth_matches <- depth %>%
  filter(ref %in% top_refs)

cat("Rows in depth matching top_refs:", nrow(depth_matches), "\n")
```


## 9. Coverage plots for top supported references

<!-- ```{r fig.height=10, fig.width=14}
#if (length(top_refs) == 0) {
#  plot.new()
#  text(0.5, 0.5, "No mapped pathogen references available for plotting.")
#} else {
#  depth %>%
#    filter(ref %in% top_refs) %>%
#    ggplot(aes(x = pos, y = depth)) +
#    geom_line() +
#    facet_wrap(~ ref, scales = "free_x") +
#    labs(
#      title = "Depth across top pathogen references",
#      x = "Reference position",
#      y = "Depth"
#    ) +
#    theme_bw(base_size = 11)
#}
```
-->

```{r chunk9, fig.height=10, fig.width=14}
depth_plot_df <- depth %>%
  filter(ref %in% top_refs, !is.na(ref))

if (length(top_refs) == 0 || nrow(depth_plot_df) == 0) {
  plot.new()
  text(
    0.5, 0.5,
    "No mapped pathogen references with depth data available for plotting."
  )
} else {
  depth_plot_binned <- depth_plot_df %>%
    group_by(ref) %>%
    mutate(bin = ntile(pos, 1000)) %>%
    group_by(ref, bin) %>%
    summarise(
      pos = median(pos),
      depth = mean(depth),
      .groups = "drop"
    )

  ggplot(depth_plot_binned, aes(x = pos, y = depth)) +
    geom_line() +
    facet_wrap(~ ref, scales = "free_x") +
    labs(
      title = "Depth across top pathogen references (binned)",
      x = "Reference position",
      y = "Mean depth"
    ) +
    theme_bw(base_size = 11)
}
```

## 10. Contig evidence

```{r chunk10}
datatable(best_contig_hits, options = list(pageLength = 20, scrollX = TRUE))
```

## 11. Integrated interpretation

```{r chunk11} 
integrated_hits <- species_tbl %>%
  select(name, taxonomy_id, new_est_reads, fraction_total_reads) %>%
  left_join(
    priority_hits %>% select(name, stevia_priority, priority_match),
    by = "name"
  ) %>%
  mutate(
    stevia_priority = replace_na(stevia_priority, FALSE),
    priority_match = replace_na(priority_match, "")
  ) %>%
  arrange(desc(new_est_reads))

datatable(integrated_hits, options = list(pageLength = 20, scrollX = TRUE))
```

A pathogen candidate is strongest when it has support from multiple layers:

1. **classifier evidence** in Kraken2/Bracken,
2. **reference mapping support** with meaningful mapped reads and coverage,
3. **contig-level BLAST support** from assembled non-host reads,
4. **biological plausibility** for Stevia or plant-associated systems.

For fungi, mapping breadth can be conservative because RNA-seq reads are aligned back to genomic references rather than transcript references.

## 12. Session info

```{r chunk12}
sessionInfo()
```

